---
title: "Reproducible Research Week 2 Project 1"
output: 
      html_document:
            keep_md: true
---

## Reproducible research_Coursera R course

## Week2, Project 1

Instruction can be found here:

https://www.coursera.org/learn/reproducible-research/peer/gYyPt/course-project-1


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = "asis")
knitr.table.format = "html"
```

#### Loading the data
```{r load, results = "hide"}
setwd("C:/Users/Xing/Documents/Coursera_R/ReproResearch_Project1")
data <- read.csv("activity.csv")
```

#### Preprocessing the data
```{r preprocess, results = "hide"}
data$date <- as.Date(data$date) # Convert the date column as Date
```

### Part A. What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

#### 1. Calculate the total number of steps taken per day
```{r nr.steps}
steps_day_total <- aggregate(data$steps, by= list(data$date), sum,na.rm =TRUE)
colnames(steps_day_total)<-c("date","total steps")
``` 

#### 2. Make a histogram of the total number of steps taken each day
```{r nr.steps histogram, figure.height = 4}
hist(steps_day_total$`total steps`, breaks = 10, xlab = c("Steps"),main = c("Total number of steps taken each day"))
```

#### 3. Calculate and report the mean and median of the total number of steps taken per day
``` {r mean/median steps}
mean_steps_day_total <- mean(steps_day_total$`total steps`, na.rm=TRUE) # mean
print(mean_steps_day_total)
median_steps_day_total <- median(steps_day_total$`total steps`, na.rm=TRUE) # median
print(median_steps_day_total)
```

### Part B. What is the average daily activity pattern?
#### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
``` {r plot 5 min interval - average steps in that interval}
steps_interval <- aggregate(data$steps, by = list(data$interval), mean,na.rm =TRUE)
colnames(steps_interval)<-c("interval","steps")
plot(steps_interval$interval, steps_interval$steps, type = "l", xlab = c("5-minute interval"), ylab = c("Average number of steps"), main = c("Average daily activity pattern"))
```

#### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
``` {r max number of steps at which interval}
max(steps_interval$steps) # max number of steps
steps_interval[match(max(steps_interval$steps),steps_interval$steps),1] # max number of steps at this interval
```

### Part C. Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.
#### 1. Total number of missing step values in the dataset
``` {r number of NA values in dataset}
sum(is.na(data$steps))
```

#### 2. & 3. Filling in all of the missing values in the dataset.
Create a new dataset that is equal to the original dataset but with the missing data filled in.
Use the average number of steps for that 5-minute interval.
``` {r fill in NA values}
data_fillNA <- data # duplicate the original dataset
# fill the NA values
for (i in 1: nrow(data)) {
      if (is.na(data$steps[i])) {
            data_fillNA$steps[i] = steps_interval$steps[match(data$interval[i], steps_interval$interval)]
      }
}
# show summary of the dataset after filling in NA values
library(knitr)
kable(summary(data_fillNA))
```

####  4. Make a histogram of the total number of steps taken each day 
``` {r nr.steps histogram (using new dataset: data_fillNA), figure.height = 4}
steps_day_total_fillNA <- aggregate(data_fillNA$steps, by= list(data_fillNA$date), sum, na.rm =TRUE)
colnames(steps_day_total_fillNA)<-c("date","total steps")
hist(steps_day_total_fillNA$`total steps`, breaks = 10, xlab = c("Steps"),main = c("Total number of steps taken each day New Dataset"))
```
Calculate and report the mean and median total number of steps taken per day. 
Do these values differ from the estimates from the first part of the assignment? 
``` {r mean/median steps (using new dataset: data_fillNA), figure.height = 4}
# mean and median data using the new dataset without NA values
mean_steps_day_total_fillNA <- mean(steps_day_total_fillNA$`total steps`, na.rm=TRUE) # mean
print(mean_steps_day_total_fillNA)
median_steps_day_total_fillNA <- median(steps_day_total_fillNA$`total steps`, na.rm=TRUE) # median
print(median_steps_day_total_fillNA)

# mean and median data using the old dataset with NA values
print(mean_steps_day_total)
print(median_steps_day_total)
```

What is the impact of imputing missing data on the estimates of the total daily number of steps?
``` {r}
print("After filling missing data using the mean steps associated with the specific time interval, the mean and average values increased.")
```

### Part D. Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

#### 1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
``` {r weekday/weekend}
data_fillNA$'weekday weekend' <-weekdays(data_fillNA$date, abbreviate = TRUE)
data_fillNA$'weekday weekend'<-sub(c("S(at|un)"),0, data_fillNA$'weekday weekend')
for (i in 1:nrow(data_fillNA)){
      if (data_fillNA$'weekday weekend'[i]!= 0) {
            data_fillNA$'weekday weekend'[i]= 1
            }
}
data_fillNA$'weekday weekend'<-factor(data_fillNA$'weekday weekend', levels = c(0,1), labels = c("Weekends","Weekdays")) 
# convert the variable as factors
# 0 as weekend
# 1 as weekday
library(knitr)
kable(summary(data_fillNA))
```

#### 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
``` {r plot weekday/weekend steps}
steps_interval_week <- aggregate(data_fillNA$steps, by= list(data_fillNA$interval,data_fillNA$'weekday weekend'), mean, na.rm =TRUE)
colnames(steps_interval_week)<-c("interval","weekday_weekend","steps")
library(ggplot2)
p<-ggplot(steps_interval_week, aes(x= interval, y = steps, group = weekday_weekend, colour = weekday_weekend)) +
      geom_line() +
      labs(x ="5-min Interval", y="Averages Steps", title = "Averages steps across all weekedays/weekends")
print(p)
```
